---
title: "03_Preliminary_Analysis"
author: "Jordan Katchen"
date: "2026-02-02"
output: html_document
---

# Introduction

This file documents the preliminary exploration of our dataset, conducted before pre-registration. Therefore, this file provides important context for further analyses.

# Preparations

Loading library:
```{r}
library(ggplot2)
library(tidyverse)
library(here)
```

Clean data files are stored in a public OSF project. The code below checks whether each file already exists locally before downloading, so repeated runs do not re-fetch files unnecessarily.

OSF project: https://osf.io/kgpyd/

```{r}
# Each file: its name on OSF, and where to save it locally
files <- list(
  list(name = "Cleaned_Master_Data.rds",            path = here("data", "clean-data", "Cleaned_Master_Data.rds")),
  list(name = "DOI_Level_Data.csv", path = here("data", "clean-data", "DOI_Level_Data.csv")),
  list(name = "Author_Level_Data.csv",           path = here("data", "clean-data", "Author_Level_Data.csv"))
)

# Create the folder if it isn't there yet
dir.create(here("data", "clean-data"), recursive = TRUE, showWarnings = FALSE)

# List all files in the OSF project (no login needed if it's public)
osf_files <- osf_retrieve_node("https://osf.io/kgpyd/") |>  # <- your project ID
  osf_ls_files()

# Download anything not already saved
walk(files, ~{
  if (!file.exists(.x$path)) {
    message("Downloading: ", .x$path)
    osf_files |>
      filter(name == .x$name) |>            # pick the file we want
      # path is a folder, not a filename — the file keeps its OSF name
      osf_download(path = dirname(.x$path), conflicts = "skip")
  } else {
    message("Already exists: ", .x$path)
  }
})

rm(files, osf_files)
```

Loading data. The master dataset, the DOI-, and the author-level subsets are all created in `02_Data_Cleaning.Rmd`:
```{r}
PLOS_Data_Full  <- readRDS(here("data", "clean-data", "Cleaned_Master_Data.rds"))
doi_data        <- read_csv(here("data", "clean-data", "DOI_Level_Data.csv"))
author_data     <- read_csv(here("data", "clean-data", "Author_Level_Data.csv"))
```

# Summary Stats

Checking the number of unique authors:
```{r}
n_distinct(author_data$combined_id)
```

Checking the number of authors with multiple publications:
```{r}
table(author_data$n_publications)
```

Checking the number of authors with multiple awards:
```{r}
table(author_data$n_awards_nserc)
```
