---
title: "03_Preliminary_Analysis"
author: "Jordan Katchen"
date: "2026-02-02"
output: html_document
---

# Introduction

This file documents the preliminary exploration of our dataset, conducted before pre-registration. No exploratory analysis was performed that is not recorded here. Therefore, this file provides important context for further analyses.

# Thoughts

- Might want to start by filter to only corresponding authors. But, the issue with this is that this column is not super reliable... Might just have to rely on the fact that if an authors received a discovery or project grant that the researcher is senior enough that they should be pushing for sharing on all of there papers.

# Preparations

Loading library:
```{r}
library(ggplot2)
library(tidyverse)
library(here)
```

Loading data:
```{r}
PLOS_Data_Full <- readRDS(here("data", "clean-data", "Cleaned_Master_Data.rds"))
```

Sub-setting the master dataset into two separate datasets specific to my diffrent questions.
```{r}
# ~~ 1. DOI-level dataset ~~
# One row per DOI, keeping the largest NSERC award across all authors
doi_data <- PLOS_Data_Full |>
  group_by(DOI) |>
  summarise(
    # Keep publication-level columns (same for all rows within a DOI)
    across(c(Publication_Day, Publication_Month, Publication_Year,
             Primary_Topic_Field, Primary_Topic_Subfield,
             Corresponding_Author_Country, First_Author_Country,
             Data_Generated, Data_Shared, Data_Location,
             Code_Generated, Code_Shared, Code_Location,
             Quarter, source), first),
    # Largest raw NSERC award across all authors on this DOI
    nserc_top_award_5yr = max(nserc_top_award_5yr, na.rm = TRUE),
    nserc_top_year_5yr  = nserc_top_year_5yr[which.max(nserc_top_award_5yr)],
    # Highest NSERC z-score across all authors on this DOI (may reflect a different author)
    nserc_top_z_5yr     = max(nserc_top_z_5yr, na.rm = TRUE)
  ) |>
  # max() returns -Inf for all-NA groups; convert back to NA
  mutate(
    nserc_top_award_5yr = na_if(nserc_top_award_5yr, -Inf),
    nserc_top_z_5yr     = na_if(nserc_top_z_5yr,     -Inf)
  ) |>
  ungroup()

# ~~ 2. Author-level dataset ~~
# One row per author, summarising data and code sharing across all publications
author_data <- PLOS_Data_Full |>
  group_by(combined_id) |>
  summarise(
    n_publications = n_distinct(DOI),
    # Data sharing
    n_data_generated = sum(Data_Generated == TRUE, na.rm = TRUE),
    n_data_shared    = sum(Data_Shared == TRUE,    na.rm = TRUE),
    n_data_linked    = sum(str_detect(Data_Location, regex("online", ignore_case = TRUE)), na.rm = TRUE),
    data_locations   = paste(sort(unique(na.omit(Data_Location))), collapse = "; "),
    # Code sharing
    n_code_generated = sum(Code_Generated == TRUE, na.rm = TRUE),
    n_code_shared    = sum(Code_Shared == TRUE,    na.rm = TRUE),
    code_locations   = paste(sort(unique(na.omit(Code_Location))), collapse = "; "),
    # Funding info (author-level columns are identical across rows, so first() is safe)
    nserc_top_award_all = first(nserc_top_award_all),
    nserc_top_year_all  = first(nserc_top_year_all),
    nserc_top_z_all     = first(nserc_top_z_all),
    n_awards_nserc      = first(n_awards_nserc)
  ) |>
  ungroup()
```

Subsetting to only authors with an award.
```{r}
author_data_sub <- author_data |>
  filter(!is.na(nserc_top_z_all))
```

# Exploration

```{r}
# Classify each DOI by both matched grants and CrossRef funders
funding_comparison <- PLOS_Data_Full |>
  group_by(DOI) |>
  summarise(
    # From matched grant data
    has_nserc_matched  = any(!is.na(nserc_top_award_5yr)),
    has_cihr_matched   = any(!is.na(cihr_top_award_5yr)),
    # From CrossRef funders string
    has_nserc_crossref = any(str_detect(funders, regex("nserc|natural sciences and engineering|10\\.13039/501100000015", ignore_case = TRUE)), na.rm = TRUE),
    has_cihr_crossref  = any(str_detect(funders, regex("cihr|canadian institutes of health research|10\\.13039/501100000024", ignore_case = TRUE)), na.rm = TRUE)
  ) |>
  mutate(
    funding_matched = case_when(
      has_nserc_matched & has_cihr_matched ~ "NSERC & CIHR",
      has_nserc_matched                   ~ "NSERC only",
      has_cihr_matched                    ~ "CIHR only",
      TRUE                                ~ "No funding info"
    ) |> factor(levels = c("NSERC & CIHR", "NSERC only", "CIHR only", "No funding info")),
    funding_crossref = case_when(
      has_nserc_crossref & has_cihr_crossref ~ "NSERC & CIHR",
      has_nserc_crossref                    ~ "NSERC only",
      has_cihr_crossref                     ~ "CIHR only",
      TRUE                                  ~ "No funding info"
    ) |> factor(levels = c("NSERC & CIHR", "NSERC only", "CIHR only", "No funding info")),
    # Agreement: both sources agree on NSERC, CIHR, or both
    funding_agreement = case_when(
      (has_nserc_matched & has_cihr_matched) & (has_nserc_crossref & has_cihr_crossref) ~ "NSERC & CIHR",
      has_nserc_matched & has_nserc_crossref & !has_cihr_matched & !has_cihr_crossref   ~ "NSERC only",
      has_cihr_matched  & has_cihr_crossref  & !has_nserc_matched & !has_nserc_crossref ~ "CIHR only",
      TRUE                                                                               ~ "No funding info"
    ) |> factor(levels = c("NSERC & CIHR", "NSERC only", "CIHR only", "No funding info"))
  )

# Plot side by side
funding_comparison |>
  pivot_longer(
    cols      = c(funding_matched, funding_crossref, funding_agreement),
    names_to  = "source",
    values_to = "funding_status"
  ) |>
  mutate(source = recode(source,
    funding_matched   = "Matched Grants",
    funding_crossref  = "CrossRef Funders",
    funding_agreement = "Agreement"
  ) |> factor(levels = c("Matched Grants", "CrossRef Funders", "Agreement"))) |>
  count(source, funding_status) |>
  group_by(source) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = source, y = prop, fill = funding_status)) +
  geom_col() +
  geom_text(aes(label = paste0(n, "\n(", scales::percent(prop, accuracy = 0.1), ")")),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_manual(values = c(
    "NSERC & CIHR"    = "#2166ac",
    "NSERC only"      = "#4dac26",
    "CIHR only"       = "#d01c8b",
    "No funding info" = "#b8b8b8"
  )) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Funding Coverage: Matched Grants vs. CrossRef Funders vs. Agreement",
    x     = NULL,
    y     = "Share of Publications",
    fill  = NULL
  ) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

```{r}
author_data |>
  filter(!is.na(nserc_top_award_all) | !is.na(cihr_top_award_all)) |>
  count(n_publications) |>
  ggplot(aes(x = n_publications, y = n)) +
  geom_col(fill = "#2166ac") +
  geom_text(aes(label = n), vjust = -0.5, size = 3) +
  labs(
    title = "Publications per Funded Author",
    x     = "Number of Publications",
    y     = "Number of Authors"
  ) +
  theme_minimal()
```

```{r}
# Classify each author by funding status
author_funding_status <- PLOS_Data_Full |>
  group_by(combined_id) |>
  summarise(
    n_publications     = n_distinct(DOI),
    has_nserc          = any(!is.na(nserc_top_award_all)),
    has_cihr           = any(!is.na(cihr_top_award_all))
  ) |>
  mutate(
    funding_status = case_when(
      has_nserc & has_cihr ~ "NSERC & CIHR",
      has_nserc            ~ "NSERC only",
      has_cihr             ~ "CIHR only",
      TRUE                 ~ "No funding info"
) |> factor(levels = c("NSERC & CIHR", "NSERC only", "CIHR only", "No funding info"))
  )

# Panel 1: number of authors per funding status
author_funding_status |>
  count(funding_status) |>
  ggplot(aes(x = funding_status, y = n, fill = funding_status)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.5) +
  scale_fill_manual(values = c(
    "NSERC & CIHR"   = "#2166ac",
    "NSERC only"     = "#4dac26",
    "CIHR only"      = "#d01c8b",
    "No funding info" = "#b8b8b8"
  )) +
  labs(
    title = "Authors by Funding Source",
    x     = NULL,
    y     = "Number of Authors"
  ) +
  theme_minimal()

# Panel 2: publication counts per author, by funding status
author_funding_status |>
  ggplot(aes(x = n_publications, fill = funding_status)) +
  geom_histogram(binwidth = 1, colour = "white") +
  scale_fill_manual(values = c(
    "NSERC & CIHR"    = "#2166ac",
    "NSERC only"      = "#4dac26",
    "CIHR only"       = "#d01c8b",
    "No funding info" = "#b8b8b8"
  )) +
  facet_wrap(~ funding_status, scales = "free_y") +
  labs(
    title = "Publications per Author by Funding Status",
    x     = "Number of Publications",
    y     = "Number of Authors"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Publications per Author

This visualization shows how many authors have 1, 2, 3, etc. publications in the dataset. Most authors unsurprisingly only one publication.
```{r}
Plos_Data %>%
  count(combined_id, name = "n_rows") %>%  # Count publications per author
  count(n_rows) %>%  # Count how many authors have each publication count
  ggplot(aes(x = n_rows, y = n)) +
  geom_col(fill = "skyblue", color = "black") +
  geom_text(aes(label = n), vjust = -0.3) +  # Add count labels on bars
  labs(
    x = "Number of Publications",
    y = "Author Count",
    title = "Distribution of Publications per Author"
  ) +
  theme_minimal()
```

## Awards Distribution

Next, we'll create a summary dataset that we'll use across multiple visualizations. This consolidates our funding data and replaces missing values with zeros.
```{r}
funding_summary <- Plos_Data %>%
  select(combined_id, starts_with("n_awards_"), starts_with("total_amount_")) %>%
  distinct() %>%  # Get unique author records
  mutate(across(everything(), ~replace_na(., 0))) %>%  # Replace all NAs with 0
  mutate(
    total_awards = n_awards_NSERC + n_awards_CIHR,
    total_funding = total_amount_NSERC + total_amount_CIHR
  )
```

This plot examines the total number of awards (combining both NSERC and CIHR) that each author has received.
```{r}
funding_summary %>%
  count(total_awards) %>%  # Count authors with each total_awards value
  ggplot(aes(x = total_awards, y = n)) +
  geom_col(fill = "skyblue", color = "black") +
  geom_text(aes(label = n), vjust = -0.3) +
  labs(
    x = "Number of Awards",
    y = "Author Count",
    title = "Distribution of Awards per Author"
  ) +
  theme_minimal()
```

## Relationship Between Awards and Funding

These scatter plots explore whether more awards correlate with higher total funding amounts.

### NSERC Funding
```{r}
funding_summary %>%
  ggplot(aes(x = n_awards_NSERC, y = total_amount_NSERC)) +
  geom_jitter(colour = "black", size = 1, shape = 1, width = 0.1) +
  labs(
    x = "Number of NSERC Awards",
    y = "Total NSERC Funding ($)",
    title = "NSERC: Awards vs. Total Funding"
  ) +
  theme_minimal()
```

### CIHR Funding
```{r}
funding_summary %>%
  ggplot(aes(x = n_awards_CIHR, y = total_amount_CIHR)) +
  geom_jitter(colour = "black", size = 1, shape = 1, width = 0.1) +
  scale_x_continuous(breaks = scales::breaks_pretty()) +
  labs(
    x = "Number of CIHR Awards",
    y = "Total CIHR Funding ($)",
    title = "CIHR: Awards vs. Total Funding"
  ) +
  theme_minimal()
```

### Combined Agency Comparison

This plot overlays both NSERC and CIHR data to compare patterns between the two funding agencies.
````{r}
funding_summary %>%
  tidyr::pivot_longer(
    cols = -combined_id,
    names_to = c(".value", "agency"),
    names_pattern = "(.+)_(NSERC|CIHR)"
  ) %>%
  filter(n_awards > 0) %>%  # Remove rows with no awards
  ggplot(aes(x = n_awards, y = total_amount, color = agency)) +
  geom_jitter(size = 1, shape = 1, width = 0.1) +
  scale_color_manual(
    values = c("NSERC" = "#E69F00", "CIHR" = "#56B4E9"),
    labels = c("CIHR", "NSERC")
  ) +
  labs(
    x = "Number of Awards",
    y = "Total Award Amount ($)",
    color = "Funding Agency",
    title = "Comparison of Awards vs. Funding Amounts by Agency"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
````
